Extreme climatic events with asymmetric tail-dependence affect population-extinction risk

### 2019-07-16 MainText.Rmd
edits, still suppmat fig caption short need to be added

\textbf{\textit{Keywords:}} extinction risk, extreme events, tail dependence, copula, density-dependence model

@@

<!--How extinction risk is related to extreme events : past studies
consequences of extreme events on population life history, community-->
Earlier studies showed ECEs may affect species abundances (species re-ordering) through 
life-history traits and this alteration in community structures could cascade 
through higher hierarchical levels and lead to local extinction of 
species [@thomas2004extinction; @edgar2010nino; @moreno2011extreme]. For example, 
a study based on annal survival data showed changes
in prey communities induced by extreme flooding affected predators in 
aquatic and adjacent riparian habitats [@chiu2013impact]. Similarly, 
$43$ years of ring-recovery data was used to examine the demographic impact of ECEs 
through temporal variability in juvenile, immature and adult survival
of European shags on the Isle of May in eastern Scotland [@frederiksen2008demographic].
A recent study [@mcdermott2017sensitivity] investigated the sensitivity 
of $41$ UK butterfly species to four different ECEs (extreme heat, extreme cold, 
extreme precipitation, drought) over $37$ years (1976-2012) at the site level, across 
each species' life-stages. Successive ECEs with short time intervals 
could weaken the resistance of plant communities too [@dreesen2014successive]. 
All these studies indicated that environmental stochasticity (through ECEs) 
can lead to demographic stochasticity (through influence on reproductive or 
mortality rate) of different age class of a population making it vulnerable 
to local extinction.

<!--How copula could be relevent in estimating extreme event's 
influence on extinction risk : introduce Taildep. concept-->
A recent study [@Ghosh_copula] showed the importance, causes 
and consequences of copula structures in associations between ecologically
important variables and demonstrated one example of density-independent metapopulation model where 
copula was used to estimate extinction risk. Copula, a rank based measure of association, is much more informative 
than linear correlation measure and a widely applied tool in dependence 
modeling [@joe2014_dependence; @nelsen2006_copula]. We here introduce the concept of copula in a
manner that follows [@Ghosh_copula]. Copulas separate the information content of a bivariate dataset,
$(x_n, y_n)$ for $n = 1 \ldots N$, into two non-overlapping parts: the information in the marginal 
distributions (which is not about the association between the variables) and the rest of 
the information (which is solely about the association). The association information is revealed by 
the copula plot ($v_n$ against $u_n$) of the raw variables ($y_n$ vs. $x_n$) in Fig...., where 
$u_n$ is the rank of $x_n$ in the set ${x_1, x_2, \ldots , x_n }$, divided by $N + 1$; 
and $v_n$ is the rank of $y_n$ in the set ${y_1, y_2, \ldots , y_n }$, also divided
by $N + 1$. Here the rank of the smallest element of each sets is considered to be 1 and the rank of
the largest element is $N$. We refer to the $u_n$ and $v_n$ as normalized ranks of the $x_n$ and $y_n$. 
Ranking makes the marginal distributions uniform, isolating only the information on 
association between the variables. Here, we will extend the application of copula 
in estimating extinction risk as introduced by [@Ghosh_copula] on why the potential use of copula
in modelling extinction risk goes beyond the density-independent model 
and even applicable for a simple age-structured population where its vital rates are influenced by ECEs.

<!--tail dep. of copula & extinction risk-->
A major benefit of the copula approach is that it can effectively capture tail dependence. 
Tail dependence, already exemplified in Fig..... (add Fig.), is associations between the more extreme 
values of distributions. For two positively associated variables, left-tail 
(respectively, right-tail) dependence is stronger association between
values in the left or lower portions, i. e. smaller values 
(respectively, the right or upper portions, i.e. higher values) 
of the two distributions. For example, extreme cold winter during growing season
might reduce the adult reproduction rate as well as larval survival (i.e. stronger 
tail dependence between vital rates when temperature is too cold than normal), which 
could lead to population crash and thereby inflate the extinction risk. Similarly, in context of 
density-dependent metapopulation models, if extreme environmental fluctuations across sites 
appear to be as synchronous (left tail dependent or right tail dependent between patches 
with same correlation as in Fig.....) then it could wipe out some metapopulation patches 
during unfavorable years unless they are rescued by the nearby populations. This scenario 
could be more complex depending 
on the dynamics of the model. Though tail-dependence 
property of a copula has been used for risk-analysis in other fields like finance, 
hydrology etc. [@alexander2008market; @li2013; @chen2011flood; @Goswami2018], but to our knowledge
copula have never been used in estimating the effect of ECEs through extinction risk of age structured 
model or density-dependent metapopulations.

@@

<!--test hypothesis from Q3 : switching of ext risk with density-dependence metapopulation model with d=0-->
Given hypothesis from (Q2) was valid which scrutinized that whereas 
left tail dependence in noise made extinction risk 
higher in undercompensatory regime ($r < r_c$), it was the 
right tail dependent noise
which appeared to evoke higher extinction risk for metapopulations
in overcompensatory regime ($r > r_c$), we tested our third hypothesis for (Q3).
We considered
the switching to take place near $r_c$, not exactly at $r = r_c$ because the transition point $r_c$
from under to overcompensatory dynamics was expected to differ in stochastic environment than its 
deterministic analog depending on the underlying nonlinearity of the model.
We generated plots with extinction risk along y-axis
for both left tail and right tail dependent noises 
against the varying range of $r$ when no dispersal ($d=0$) was considered.
We excluded the possibility of dispersal in this case to avoid any rescue effect among patches. 
For each model with extreme left and extreme right tail dependent noises, we calculated 
2.5$^{th}$ and 97.5$^{th}$ quantiles
of 10000 simulations with the probability of computed
extinction risk at each $r$. We compared $r_c$ and the value of $r$ at which 
95\% CI of extinction risk from both noises overlapped
with each other. This could give us an idea for supporting our hypothesis.

# Results

@@

We found density-dependence metapopulation models (with $n=5$ patches) in noisy 
environment showed similar patterns for extinction risk as density-independent 
metapopulation model showed in earlier studies [@Ghosh_copula]. For each density-dependent 
model listed in Table \ref{tab_model_summary}, we observed higher (or lower) extinction risk for 
the metapopulations when noises across multiple patches were left- (or right) tail dependent.
But that observation remained valid when the model parameter, usually growth rate $(r)$, of each model was 
sufficiently low so that under-compensatory dynamics (monotonous approach towards
its stable equilibrium) for the deterministic analog of the model was ensured. When the model 
was in over-compensatory regime (oscillatory approach towards stable equilibrium) for higher 
value of $r$, we observed right (or left) tail dependent noise caused higher (or lower) extinction 
risk for the metapopulations. Figs. \ref{fig_extrisk_allmodel_localdisp} shows variation of extinction risk 
against local dispersal rate ($d$) for six density dependent metapopulation models 
where growth rate $r$ was taken either from under-compensatory or from over-compensatory 
regime. We also got similarly consisent results with 
our prediction (see Q2 from \nameref{intro}) for global 
dispersal (Fig. \ref{SM-fig_extrisk_allmodel_globaldisp}). 
Figs. (\ref{SM-fig_ricker_risk_local_disp_scl_1} - \ref{SM-fig_risk_local_disp_abrewer_r_vary_scl_0.6}) 
shows how extinction risk changed against local dispersal rate for a whole range
of $r$, whereas Figs. (\ref{SM-fig_ricker_risk_global_disp_scl_1} - 
\ref{SM-fig_risk_global_disp_abrewer_r_vary_scl_0.6}) 
represents the same but when global dispersal took place between patches. 
Both case showed the signature of 
"rescue-effect" (i.e. by lowering extinction risk) when there was 
a finite dispersal compared to the one with 
$d=0$.

<!--Q3 : metapopulation results : switching near rc for d=0, Fig. 3-->
Following Figs. (\ref{SM-fig_ricker_risk_local_disp_scl_1} - 
\ref{SM-fig_risk_global_disp_abrewer_r_vary_scl_0.6}) we hypothesized the switching 
between the extinction risk generated by left-tail dependent and right-tail dependent noise 
occurred near the critical parameter $r_c$ (see Q3). For simplicity we only consider 
$d=0$ case, i.e., when there was no dispersal between patches (see \nameref{methods}). 
Fig. \ref{fig_extrisk_d0} shows higher extinction risk with left-tail dependent noise 
for $r<r_c$ and with right-tail dependent noise for $r>r_c$. $r_c$ was the vertical
line in Fig. \ref{fig_extrisk_d0} indicating the value of $r$ for each model where 
their deterministic analog switches from under- to over-compensatory dynamics. 
Though the switching visually appered far from the theoretical $r_c$ value for Verhulst 
and Austin-Brewer model, they were consistent with our hypothesis as they had comperatively 
smaller range of $r$ than the other models.

>>>>

\textbf{\textit{Keywords:}} extinction risk, extreme events, tail dependence, copula, density-dependent model

@@

<!--How extinction risk is related to extreme events : past studies
consequences of extreme events on population life history, community-->
Earlier studies showed ECEs may affect species abundances (species re-ordering) through 
life-history traits and this alteration in community structures could cascade 
through higher hierarchical levels and lead to local extinction of 
species [@thomas2004extinction; @edgar2010nino; @moreno2011extreme]. 
<!--Should I mention idea of match-mismatch hypothesis?-->
For example, 
a study based on annal survival data showed changes
in prey communities induced by extreme flooding affected predators in 
aquatic and adjacent riparian habitats [@chiu2013impact] via trophic linkages. Similarly, 
$43$ years of ring-recovery data was used to examine the demographic impact of ECEs 
through temporal variability in juvenile, immature and adult survival
of European shags on the Isle of May in eastern Scotland [@frederiksen2008demographic].
A recent study [@mcdermott2017sensitivity] investigated the sensitivity 
of $41$ UK butterfly species to four different ECEs (extreme heat, extreme cold, 
extreme precipitation, drought) over $37$ years (1976-2012) at the site level, across 
each species' life-stages. Experimental study showed successive ECEs with short time intervals 
could weaken the resistance of plant communities too [@dreesen2014successive]. 
All these studies indicated that environmental stochasticity (through ECEs) 
can lead to demographic stochasticity (through influence on reproductive or 
mortality rate) of different age class of a population making it vulnerable 
to local extinction.

<!--How copula could be relevent in estimating extreme event's 
influence on extinction risk : introduce Taildep. concept-->
A recent study [@Ghosh_copula] showed the importance, causes 
and consequences of copula structures in associations between ecologically
important variables and demonstrated one example of density-independent metapopulation model where 
copula was used to estimate extinction risk. Copula, a rank based measure of association, is much more informative 
than linear correlation measure and a widely applied tool in dependence 
modeling [@joe2014_dependence; @nelsen2006_copula]. We here introduce the concept of copula in a
manner that follows @Ghosh_copula. Copulas separate the information content of a bivariate dataset,
$(x_n, y_n)$ for $n = 1 \ldots N$, into two non-overlapping parts: the information in the marginal 
distributions (which is not about the association between the variables) and the rest of 
the information (which is solely about the association). The association information is revealed by 
the copula plot ($v_n$ against $u_n$) of the raw variables ($y_n$ vs. $x_n$) in Fig...., where 
$u_n$ is the rank of $x_n$ in the set ${x_1, x_2, \ldots , x_n }$, divided by $N + 1$; 
and $v_n$ is the rank of $y_n$ in the set ${y_1, y_2, \ldots , y_n }$, also divided
by $N + 1$. Here the rank of the smallest element of each sets is considered to be 1 and the rank of
the largest element is $N$. We refer to the $u_n$ and $v_n$ as normalized ranks of the $x_n$ and $y_n$. 
Ranking makes the marginal distributions uniform, isolating only the information on 
association between the variables. Here, we will extend the application of copula 
in estimating extinction risk as introduced by [@Ghosh_copula] on why the potential use of copula
in modelling extinction risk goes beyond the density-independent model 
and even applicable for a simple age-structured population where its vital rates are influenced by ECEs.

<!--tail dep. of copula & extinction risk-->
A major benefit of the copula approach is that it can effectively capture the tail dependence. 
Tail dependence, already exemplified in Fig..... (add Fig.), is association between the more extreme 
values of distributions. For two positively associated variables, left-tail 
(respectively, right-tail) dependence is stronger association between
values in the left or lower portions, i. e. for smaller values 
(respectively, the right or upper portions, i.e. for higher values) 
of the two distributions. Stronger dependence in either tail gives rise to asymmetric copula 
structure for the joint distribution of two variables. 
For example, extreme cold winter during growing season
might reduce the adult reproduction rate as well as larval survival (i.e. stronger 
tail dependence between vital rates when temperature is too cold than normal), which 
could lead to population crash and thereby inflate the extinction risk. Similarly, in context of 
density-dependent metapopulation models, if extreme environmental fluctuations across sites 
appear to be as synchronous (left tail dependent or right tail dependent between patches 
with same correlation as in Fig.....) then it could wipe out some metapopulation patches 
during unfavorable years unless they are rescued by the nearby populations. This scenario 
could be more complex depending 
on the dynamics of the model (see \nameref{results}). Though tail-dependence 
property of a copula has been used for risk-analysis in other fields like finance, 
hydrology etc. [@alexander2008market; @li2013; @chen2011flood; @Goswami2018], but to our knowledge
copula have never been used in estimating the effect of ECEs through extinction risk of age structured 
model or density-dependent metapopulations.

@@

<!--test hypothesis from Q3 : switching of ext risk with density-dependence metapopulation model with d=0-->
Given hypothesis from (Q2) was valid which scrutinized that whereas 
left tail dependence in noise made extinction risk 
higher in undercompensatory regime ($r < r_c$), it was the 
right tail dependent noise
which appeared to evoke higher extinction risk for metapopulations
in overcompensatory regime ($r > r_c$), we tested our third hypothesis for (Q3).
We considered
the switching to take place near $r_c$, not exactly at $r = r_c$ because the transition point $r_c$
from under to overcompensatory dynamics was expected to differ in stochastic environment than its 
deterministic analog depending on the underlying nonlinearity of the model.
We generated plots with extinction risk along y-axis
for both left tail and right tail dependent noises 
against the varying range of $r$ (along x-axis) when no dispersal ($d=0$) was considered.
We excluded the possibility of dispersal in this case to avoid any rescue effect among patches. 
For each model with extreme left and extreme right tail dependent noises, we calculated 
2.5$^{th}$ and 97.5$^{th}$ quantiles
of 10000 simulations with the probability of computed
extinction risk at each $r$. We compared $r_c$ and the value of $r$ at which 
95\% CI of extinction risk from both noises overlapped
with each other. This could give us an idea for supporting our hypothesis.

# Results \label{results}

@@

We found density-dependence metapopulation models (with $n=5$ patches) in noisy 
environment showed similar patterns for extinction risk as density-independent 
metapopulation model showed in earlier studies [@Ghosh_copula]. For each density-dependent 
model listed in Table \ref{tab_model_summary}, we observed higher (or lower) extinction risk for 
the metapopulations when noises across multiple patches were left- (or right) tail dependent.
But that observation remained valid when the model parameter, usually growth rate $(r)$, of each model was 
sufficiently low so that under-compensatory dynamics (monotonous approach towards
its stable equilibrium) for the deterministic analog of the model was ensured. When the model 
was in over-compensatory regime (oscillatory approach towards stable equilibrium) for higher 
value of $r$, we observed right (or left) tail dependent noise caused higher (or lower) extinction 
risk for the metapopulations. Fig. \ref{fig_extrisk_allmodel_localdisp} shows variation of extinction risk 
against local dispersal rate ($d$) for six density dependent metapopulation models 
where growth rate $r$ was taken either from under-compensatory or from over-compensatory 
regime. We also got similarly consisent results with 
our prediction (see Q2 from \nameref{intro}) for global 
dispersal (Fig. \ref{SM-fig_extrisk_allmodel_globaldisp}). 
Figs. (\ref{SM-fig_ricker_risk_local_disp_scl_1} - \ref{SM-fig_risk_local_disp_abrewer_r_vary_scl_0.6}) 
shows how extinction risk changed against local dispersal rate for a whole range
of $r$, whereas Figs. (\ref{SM-fig_ricker_risk_global_disp_scl_1} - 
\ref{SM-fig_risk_global_disp_abrewer_r_vary_scl_0.6}) 
represents the same but for the case when global dispersal took place between patches. 
Both case showed the signature of 
"rescue-effect" (i.e. by lowering extinction risk) when there was 
a finite dispersal compared to the one with 
$d=0$.

<!--Q3 : metapopulation results : switching near rc for d=0, Fig. 3-->
Following Figs. (\ref{SM-fig_ricker_risk_local_disp_scl_1} - 
\ref{SM-fig_risk_global_disp_abrewer_r_vary_scl_0.6}) we hypothesized that the switching 
between the extinction risk generated by left-tail dependent and right-tail dependent noise 
occurred near the critical parameter $r_c$ (see Q3). For simplicity we only consider 
$d=0$ case, i.e., when there was no dispersal between patches (see \nameref{methods}). 
Fig. \ref{fig_extrisk_d0} shows higher extinction risk with left-tail dependent noise 
for $r<r_c$ and with right-tail dependent noise for $r>r_c$. $r_c$ was the vertical
line in Fig. \ref{fig_extrisk_d0} indicating the value of $r$ for each model where 
their deterministic analog switches from under- to over-compensatory dynamics. 
Though the switching visually appered far from the theoretical $r_c$ value for Verhulst 
and Austin-Brewer model, they were consistent with our hypothesis as they had comperatively 
smaller range of $r$ than the other models.
